---
title: "Multiple linear regression"
subtitle: "Types of predictors"
author: "Prof. Maria Tackett"
date: "02-03-2026"
date-format: "MMMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 08-predictor-types-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
filters: 
  - webr
webr: 
  packages: ['tidyverse', 'broom', 'palmerpenguins']
---

## Announcements {.midi}

-   Lab 02 due **Thursday, February 5** at 11:59pm
-   HW 02 due Tuesday, February 10 at 11:59pm
    -   Released after class today
-   [Statistics experience](../hw/stats-experience.html) due April 21
-   SSMU Mini DataFest - February 8
    -   See [Ed Discussion](https://edstem.org/us/courses/91299/discussion/7577398) for announcement

## Computational setup

```{r packages}
#| echo: true
#| message: false

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(patchwork)
library(knitr)
library(kableExtra)
library(colorblindr)
library(palmerpenguins)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 16))
```

```{webr-r}
#| context: setup

penguins <- palmerpenguins::penguins |>
  select(body_mass_g, flipper_length_mm, bill_length_mm, species) |>
  drop_na() |>
  mutate(
    adelie = if_else(species == "Adelie", 1, 0),
    chinstrap = if_else(species == "Chinstrap", 1, 0),
    gentoo = if_else(species == "Gentoo", 1, 0)
  )
```

## Topics

-   Include indicator variables for categorical predictors in the regression model

-   Mean-center and standardize quantitative predictors

-   Use interaction terms in the model to capture differences in a predictor's effect

## Data: Palmer penguins {.midi}

The `penguins` data set contains data for penguins found on three islands in the Palmer Archipelago, Antarctica. Data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pal.lternet.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/). These data can be found in the **palmerpenguins** R package.

. . .

```{r}
#| echo: false

penguins <- penguins |>
  select(body_mass_g, flipper_length_mm, bill_length_mm, species) |>
  drop_na()

penguins
```

## Variables

**Predictors**:

-   `bill_length_mm`: Bill length in millimeters
-   `flipper_length_mm`: Flipper length in millimeters
-   `species`: Adelie, Gentoo, or Chinstrap species

**Response**: `body_mass_g`: Body mass in grams

<br>

**The goal of this analysis is to use the bill length, flipper length, and species to predict body mass.**

## Response: `body_mass_g`

```{r}
#| echo: false
ggplot(penguins, aes(x = body_mass_g)) +
  geom_density(fill = "steelblue") +
  labs(x = "Body mass in grams",
    title = "Distribution of body mass of Palmer Penguins")
```

```{r}
#| echo: false
penguins |>
  summarise(
    min = min(body_mass_g),
    median = median(body_mass_g),
    max = max(body_mass_g),
    iqr = IQR(body_mass_g)
  ) |>
  kable()
```

## Predictors {.small}

```{r}
#| echo: false
#| out.width: "100%"
p1 <- ggplot(penguins, aes(y = species)) +
  geom_bar(fill = "steelblue", color = "black") +
  labs(title = "Species", 
       y = "")

p2 <- ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(fill = "steelblue", color = "black") +
  labs(title = "",
       x = "Flipper length in mm")

p3 <- ggplot(penguins, aes(x = bill_length_mm)) +
  geom_histogram(fill = "steelblue", color = "black") +
  labs(title = "",
       x = "Bill length in mm")

p1 + p2 / p3
```

## Response vs. predictors {.small}

```{r}
#| echo: false
p4 <- ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    y = "Body mass in grams",
    x = "Species"
  )

p5 <- ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Flipper length in mm"
  )


p6 <- ggplot(penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point(color = "steelblue") +
  labs(
    y = NULL,
    x = "Bill length in mm"
  )

p4 + p5 / p6
```

## Multiple linear regression model

```{r}
#| echo: true

penguin_fit <- lm(body_mass_g ~ flipper_length_mm + species + 
                bill_length_mm, data = penguins)

tidy(penguin_fit) |>
  kable(digits = 3)
```

## Model equation

$$
\begin{align}\widehat{\text{body_mass_g}} =  -3904.387 &+27.429 \times \text{flipper_length_mm}\\
& -748.562 \times \text{Chinstrap}\\  
&+ 90.435 \times \text{Gentoo}\\
&+ 61.736 \times \text{bill_length_mm}
\end{align}
$$

::: callout-note
We will talk about why there are two terms in the model for `species` shortly!
:::

## Interpreting $\hat{\beta}_j$

-   The estimated coefficient $\hat{\beta}_j$ is the expected change in the mean of $Y$ when $X_j$ increases by one unit, <i>holding the values of all other predictor variables constant</i>.

<br>

-   **Example:** The estimated coefficient for `flipper_length_mm` is 27.429. This means for each additional millimeter in a penguin's flipper length, its body mass is expected to be greater by 27.429 grams, on average, holding species and bill length constant.

## Prediction

::: question
What is the predicted body mass for a Gentoo penguin with a flipper length of 200 millimeters and bill length of 45 millimeters?
:::

<br>

```{r}
#| echo: true

new_penguin <- tibble(
  flipper_length_mm  = 200, 
  species = "Gentoo", 
  bill_length_mm = 45
)

predict(penguin_fit, new_penguin)
```

## Intervals for prediction {.midi}

90% confidence interval for **estimated mean body mass** of Gentoo penguins with flipper length of 200 mm and bill length of 45 mm.

```{r}
#| echo: true

predict(penguin_fit, new_penguin, interval = "confidence", 
        level = 0.90) |> kable(digits = 3)
```

<br>

90% prediction interval for **estimated body mass of an individual** Gentoo penguin with flipper length of 200 mm and bill length of 45 mm.

```{r}
#| echo: true

predict(penguin_fit, new_penguin, interval = "prediction", 
        level = 0.90) |> kable(digits = 3)
```

# Categorical predictors

## Indicator variables

-   Suppose there is a categorical variable with $k$ categories (levels)

-   We can make $k$ indicator variables - one indicator for each category

-   An **indicator variable** (also called dummy variables) takes values 1 or 0

    -   1 if the observation belongs to that category
    -   0 if the observation does not belong to that category

## Indicator variables for `species`

```{r}
#| echo: true

penguins <- penguins |>
  mutate(
    adelie = if_else(species == "Adelie", 1, 0),
    chinstrap = if_else(species == "Chinstrap", 1, 0),
    gentoo = if_else(species == "Gentoo", 1, 0)
  )
```

. . .

```{r}
#| echo: false
penguins |>
  select(species, adelie, chinstrap, gentoo) |>
  slice(1, 152, 283)
```

## Indicators in the model {.midi}

-   We will use $k-1$ of the indicator variables in the model.
-   The **baseline** is the category that doesn't have a term in the model. This is also called the *reference level.*
-   The coefficients of the indicator variables in the model are interpreted as the expected change in the response compared to the baseline, holding all other variables constant.

. . .

```{r}
penguins |>
  select(species, chinstrap, gentoo) |>
  slice(1, 152, 283) # select example observations
```

## Using all $k$ indicators in R model {.midi}

::: question
Run the code below to fit a model using `flipper_length_mm`, `bill_length_mm`, and all the indicator variables for `species` to predict `body_mass_g`. What do you notice about the model output? Why did this happen?
:::

```{webr-r}
body_mass_fit <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + 
                      adelie + chinstrap + gentoo,
                    data = penguins)

tidy(body_mass_fit)
```

## Interpreting `species` {.midi}

```{r}
#| echo: false
tidy(penguin_fit, conf.int  = T) |>
  kable(digits = 3) |>
  row_spec(c(3,4), background = "#dce5b2")
```

. . .

::: incremental
-   The baseline category is `Adelie`.
-   Penguins from the Chinstrap species are expected to have a body mass that is 748.562 grams less, on average, compared to penguins from the Adelie species, holding flipper length and bill length constant.
:::

. . .

::: question
Interpret the coefficient of `Gentoo` in the context of the data.
:::

# Transforming quantitative predictors

## Centering {.midi}

-   Centering a quantitative predictor means shifting every value by some constant $C$

$$
X_{cent} = X  - C
$$

-   One common type of centering is **mean-centering**, in which every value of a predictor is shifted by its mean

-   Only quantitative predictors are centered

-   Center all quantitative predictors in the model for ease of interpretation, when using mean-centering

::: question
What is one reason we might want to center the quantitative predictors? What are the units of centered variables?
:::

## Centering {.midi}

Use the `scale()` function with `center = TRUE` and `scale = FALSE` to mean-center variables

```{r}
#| output-location: fragment 

penguins <- penguins |>
  mutate(flipper_length_cent = scale(flipper_length_mm, center = TRUE, scale = FALSE), 
         bill_length_cent = scale(bill_length_mm, center = TRUE, scale = FALSE))
```

## Original vs. mean-centered variable

::::: columns
::: {.column width="50%"}
**Original variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Flipper length in mm")
```

<br>

```{r}
#| echo: false
penguins |>
  summarise(Mean = mean(flipper_length_mm),
            SD = sd(flipper_length_mm)) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
**Mean-centered variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_cent)) +
  geom_density(fill = "darkcyan") +
  labs(x = NULL, title = "Mean-centered Flipper length in mm")
```

<br>

```{r}
#| echo: false 

penguins |>
  summarise(Mean = mean(flipper_length_cent),
            SD = sd(flipper_length_cent)) |>
  kable(digits = 3)
```
:::
:::::

<br>

. . .

## Using mean-centered variables in the model

::: question
How do you expect the model to change if we use `flipper_length_cent` and `bill_length_cent` in the model?
:::

. . .

```{r}
#| echo: false
penguin_cent_fit <- lm(body_mass_g ~ flipper_length_cent + species +
                    bill_length_cent, data = penguins)

tidy(penguin_cent_fit, conf.int = T) |>
  kable(digits = 3)
```

## Original vs. mean-centered model {.midi}

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_cent_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

::: question
What changed? What is the same?
:::

## Interpret coefficient of mean-centered predictors

::::: panel-tabset
## Question

:::: midi
::: appex
Interpret the coefficient of `bill_length_cent` in the context of the data.
:::

🔗 <https://forms.office.com/r/Cn8pp2r32U>
::::

## Submit

```{=html}
<center><iframe width="640px" height="480px" src="https://forms.office.com/r/Cn8pp2r32U?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe></center>
```
:::::

## Standardizing {.midi}

-   Standardizing a quantitative predictor mean shifting every value by the mean and dividing by the standard deviation of that variable

$$
X_{std} = \frac{X - \bar{X}}{s_X}
$$

-   Only quantitative predictors are standardized

-   Standardize all quantitative predictors in the model for ease of interpretation

::: question
What is one reason we might want to standardize the quantitative predictors? What are the units of standardized variables?
:::

## Standardizing

Use the `scale()` function with `center = TRUE` and `scale = TRUE` to standardized variables

<br>

```{r}
#| output-location: fragment 
#| 
penguins <- penguins |>
  mutate(flipper_length_std = scale(flipper_length_mm, center = TRUE, scale = TRUE), 
         bill_length_std = scale(bill_length_mm, center = TRUE, scale = TRUE))
```

## Original vs. standardized variable

::::: columns
::: {.column width="50%"}
**Original variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_density(fill = "steelblue") +
  labs(x = NULL, title = "Flipper length in mm")
```

<br>

```{r}
#| echo: false

penguins |>
  summarise(Mean = mean(flipper_length_mm),
            SD = sd(flipper_length_mm)) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
**Standardized variable**

```{r}
#| echo: false 
ggplot(penguins, aes(x = flipper_length_std)) +
  geom_density(fill = "darkcyan") +
  labs(x = NULL, title = "Standardized Flipper length in mm")
```

<br>

```{r}
#| echo: false

penguins |>
  summarise(Mean = mean(flipper_length_std),
            SD = sd(flipper_length_std)) |>
  kable(digits = 3)
```
:::
:::::

<br>

. . .

## Using standardized variables in the model

::: question
How do you expect the model to change if we use `flipper_length_std` and `bill_length_std` in the model?
:::

. . .

```{r}
#| echo: false
penguin_std_fit <- lm(body_mass_g ~ flipper_length_std + species +
                    bill_length_std, data = penguins)

tidy(penguin_std_fit, conf.int = T) |>
  kable(digits = 3)
```

## Original vs. standardized model

::::: columns
::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
tidy(penguin_std_fit) |>
  select(term, estimate) |>
  kable(digits = 3)
```
:::
:::::

::: question
What changed? What is the same?
:::

# Interaction terms

## Interaction terms

-   Sometimes the relationship between a predictor variable and the response depends on the value of another predictor variable.
-   This is an **interaction effect**.
-   To account for this, we can include **interaction terms** in the model.

## Bill length versus species

If the lines are not parallel, there is indication of a potential **interaction effect**, i.e., the slope of bill length may differ based on the species.

```{r}
#| echo: false

p1 <- ggplot(penguins, 
             aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Bill length in millimeters",
    y = "Body mass in grams"
  )

p2 <- ggplot(penguins, 
             aes(x = bill_length_mm, y = body_mass_g,
                 color = species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Bill length in millimeters", y = NULL, color = NULL) +
  theme(legend.position = c(0.3, 0.9)) +
  scale_color_OkabeIto()

p1 + p2 +
  plot_annotation(title = "Body mass versus bill length")
```

## Interaction terms in R {.midi}

::: question
Fit the model predicting `body_mass_g` using `flipper_length_mm`, `bill_length_mm`, `speices`, and the interaction between `species` and `bill_length_mm`.

-   Write the estimated regression equation for `Adelie` penguins.

-   Write the estimated regression equation for `Chinstrap` penguins.
:::

```{webr-r}

# code to fit and display model with interaction
```

## Interaction term in model {.smaller}

```{r}
#| echo: true
penguin_int_fit <- lm(body_mass_g ~ flipper_length_mm + species + bill_length_mm + species * bill_length_mm,
      data = penguins)
```

```{r}
#| echo: false
tidy(penguin_int_fit) |>
  kable(digits = 3) |>
  row_spec(c(6,7), background = "#dce5b2")
```

## Interpreting interaction terms

-   What the interaction means: The effect of bill length on the body mass is 41.035 less when the penguin is from the Chinstrap species compared to the effect for the Adelie species, holding all else constant.
-   Interpreting `bill_length_mm` for Chinstrap: For each additional millimeter in bill length, we expect the body mass of Chinstrap penguins to increase by 31.657 grams (72.692 - 41.035), holding flipper length constant.

## Summary

<br>

::: question
In general, how do

indicators for categorical predictors impact the model equation?

interaction terms impact the model equation?
:::

## Recap

-   Included indicator variables for categorical predictors in the regression model

-   Mean-centered and standardized quantitative predictors

-   Used interaction terms in the model to capture differences in a predictor's effect

## Next class

-   Inference for multiple linear regression

-   Complete [Lecture 09 prepare](../prepare/prepare-lec09.html)
