---
title: "Model conditions + diagnostics"
author: "Prof. Maria Tackett"
date: "02-10-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 12-conditions-diagnostics-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements - UPDATE

-   HW 02

-   Lab 03

-   Lecture recordings + exam 01 practice problems on menu of course website

## Review: AE 06

::: appex
📋 [sta210-sp26.github.io/ae/ae-06-mlr-inf.html](../ae/ae-06-mlr-inf.html)
:::

## Question 

-   write question about the structure that gets preserved vs. messed up depending on how you do permutation sampling.

## Computing set up

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)  
library(tidymodels)  
library(knitr)       
library(patchwork)   
library(viridis)

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Topics

-   Model conditions

-   Influential points

-   Model diagnostics

    -   Leverage

    -   Studentized residuals

    -   Cook's Distance

## Data: Duke lemurs - UPDATE {.midi}

Today's data contains a subset of the original Duke Lemur data set available in the [TidyTuesday GitHub repo](https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-08-24/readme.md). This data includes information on lemurs age 25 months or young from who were at the Duke Lemur Center when they were measured.

**The goal of the analysis is to use the age, sex, taxon, and litter size of the lemurs to understand variability in the weight.**

```{r}
#| echo: false

lemurs <- read_csv("data/lemurs-sample-young.csv")
```

## Variables

-   `weight`: Weight: Animal weight, in grams. Weights under 500g generally to nearest 0.1-1g; Weights \>500g generally to the nearest 1-20g.

-   `taxon` : Code made as a combination of the lemur’s genus and species. Note that the genus is a broader categorization that includes lemurs from multiple species. This analysis focuses on the following taxon:

    -   `ERUF`: Eulemur rufus, commonly known as Red-fronted brown lemur
    -   `PCOQ`: Propithecus coquereli, commonly known as Coquerel’s sifaka
    -   `VRUB`: Varecia rubra, commonly known as Red ruffed lemur

-   `sex` : Sex of lemur (`M`: Male, `F`: Female)

-   `age`: Age of lemur when weight was recorded (in months)

-   `litter_size`: Total number of infants in the litter the lemur was born into (this includes the observed lemur)

## Bivariate EDA

```{r}
#| echo: false

p1 <- ggplot(data = lemurs, aes(x = age, y = weight)) + 
  geom_point() + 
  labs(x = "Age in months", 
       y = "Weight in grams", 
       title = "Weight vs. age")

p2 <- ggplot(data = lemurs, aes(x = sex, y = weight)) + 
  geom_boxplot(color = "black", fill = "steelblue") + 
  labs(x = "Sex", 
       y = "", 
       title = "Weight vs. sex")

p3 <- ggplot(data = lemurs, aes(x = taxon, y = weight)) + 
  geom_boxplot(color = "black", fill = "steelblue") + 
  labs(x = "Taxon", 
       y = "Weight in grams", 
       title = "Weight vs. taxon")


p4 <- ggplot(data = lemurs, aes(x = litter_size, y = weight)) + 
  geom_point() + 
  labs(x = "Litter size", 
       y = "", 
       title = "Weight vs. litter size")


(p1 + p2) / (p3 + p4)
```

## Fit model

```{r}
lemurs_fit <- lm(weight ~ age + sex + taxon + litter_size, data = lemurs)

tidy(lemurs_fit) |> 
  kable(digits = 3)
```

# Model conditions

## Assumptions for regression {.midi}

$$
Y|X_1, \ldots, X_p \sim N(\beta_0 + \beta_1X_1 + \dots + \beta_pX_p, \sigma_\epsilon^2)
$$

1.  **Linearity:** There is a linear relationship between the response and predictor variables.
2.  **Equal variance:** The variability about the least squares line is equal for all combinations of predictors.
3.  **Normality:** The distribution of the errors (residuals) is approximately normal.
4.  **Independence:** The errors (residuals) are independent from one another.

. . .

::: question
How do we know if these assumptions hold in our data?
:::

## Linearity {.midi}

-   Look at plot of residuals versus fitted (predicted) values.
-   Linearity is satisfied if there is no discernible pattern in the plot (i.e., points randomly scattered around $\text{residuals} = 0$

. . .

```{r}
#| echo: false
#| out-width: 65%
#| fig-align: center
lemurs_aug <- augment(lemurs_fit)

ggplot(data = lemurs_aug, aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "red") + 
  labs(x = "Fitted values", 
       y = "Residuals", 
       title = "Residuals vs. Fitted Values")
```

. . .

<center>**Linearity is** **satisfied**</center>

## Residuals vs. predictors

If we need a closer look at the linearity condition, we can plot the residuals versus each quantitative predictor.

```{r}
#| echo: false
#| fig-align: center
resid_age <- ggplot(data = lemurs_aug, aes(x = age, y = .resid)) + 
   geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "red") + 
  labs(x = "Age in months", 
       y = "Residuals", 
       title = "Residuals vs. age")

resid_litter <- ggplot(data = lemurs_aug, aes(x = litter_size, y = .resid)) + 
   geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "red") + 
  labs(x = "Litter size", 
       y = "Residuals", 
       title = "Residuals vs. litter_size")

resid_age + resid_litter

```

## Example: Linearity not satisfied

```{r}
#| echo: false
#| fig-align: center
#| 
# code produced by chatgpt
# https://chatgpt.com/c/67bd1e0e-7650-8002-b467-1bcab2e3e172

# Simulate data where linearity is violated
set.seed(123)
n <- 100
x <- runif(n, -2, 2)
y <- x^2 + rnorm(n, sd = 0.8)  # Increased noise to obscure quadratic relationship

data <- tibble(x, y)

# Fit a linear model
model <- lm(y ~ x, data = data)

# Get residuals and fitted values
data <- data %>% 
  mutate(fitted = fitted(model), residuals = resid(model))

# Plot residuals vs fitted values
ggplot(data, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(subtitle = "Residuals vs Fitted Values",
       title = "Linearity not satisfied",
       x = "Fitted Values", y = "Residuals") +
  theme_bw()



```

. . .

-   If linearity is not satisfied, examine the plots of residuals versus each predictor.
-   Add higher order term(s), as needed.

## Equal variance {.midi}

-   Look at plot of residuals versus fitted (predicted) values.
-   Equal variance is satisfied if the vertical spread of the points is approximately equal for all fitted values

. . .

```{r}
#| echo: false
#| out-width: 65%
#| fig-align: center
lemurs_aug <- augment(lemurs_fit)

ggplot(data = lemurs_aug, aes(x = .fitted, y = .resid)) +
  geom_point() + 
  geom_hline(yintercept = 0, linetype = 2, color = "red") + 
  labs(x = "Fitted values", 
       y = "Residuals", 
       title = "Residuals vs. fitted values")
```

. . .

<center>**Equal variance is** **satisfied**</center>

## Example: Equal variance not satisfied

```{r}
#| echo: false
#| fig-align: center


# simulation code from Chat gpt: 
# https://chatgpt.com/c/6987e565-b408-8327-bcfe-b1ab5049c5af

set.seed(42)

# Sample size
n <- 300

# Predictor
X <- runif(n, min = 0, max = 10)

# Heteroskedastic error:
# standard deviation increases with X
sigma <- 0.5 + 0.4 * X
epsilon <- rnorm(n, mean = 0, sd = sigma)

# Linear model with non-constant variance
beta_0 <- 2.0
beta_1 <- 0.5
y <- beta_0 + beta_1 * X + epsilon

# Store in a data frame
df <- data.frame(
  X = X,
  y = y,
  sigma = sigma
)

model <- lm(y ~ X, data = df)

# Get residuals and fitted values
df <- df %>% 
  mutate(fitted = fitted(model), residuals = resid(model))

# Plot residuals vs fitted values
ggplot(df, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Equal variance not satisfied",
  subtitle = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_bw()




```

. . .

-   Condition is critical for inference

-   Address violations by applying transformation on the response

## Normality {.midi}

-   Look at the distribution of the residuals
-   Normality is satisfied if the distribution is approximately unimodal and symmetric. Inference robust to violations if $n > 30$

```{r}
#| echo: false
#| fig-align: center
#| out.width: 65%

ggplot(data = lemurs_aug, aes(x = .resid)) +
  geom_histogram(fill = "steelblue", color = "black") + 
  labs(x = "Residuals", 
       title = "Distribution of residuals")
```

. . .

<center>Distribution approximately unimodal and symmetric, aside from the outlier. There are `r nrow(lemurs)` observations, so inference robust to departures.</center>

## Independence

-   We can often check the independence condition based on the context of the data and how the observations were collected.

-   If the data were collected in a particular order, examine a scatterplot of the residuals versus order in which the data were collected.

-   If data has spatial element, plot residuals on a map to examine potential spatial correlation.

-   If there are subgroups in the data, plot the residuals by subgroup

## Independence

```{r}
#| echo: false
#| fig-align: center

resid_sex <- ggplot(data = lemurs_aug, aes(x = sex, y = .resid)) + 
  geom_boxplot(fill = "steelblue", color = "black") + 
  geom_hline(yintercept = 0, color = "red", linetype = 2) + 
  labs(x = "Sex", 
       y = "Residuals", 
       title = "Residuals vs. sex")

resid_taxon <- ggplot(data = lemurs_aug, aes(x = sex, y = .resid)) + 
  geom_boxplot(fill = "steelblue", color = "black") + 
  geom_hline(yintercept = 0, color = "red", linetype = 2) + 
  labs(x = "Taxon", 
       y = "Residuals", 
       title = "Residuals vs. taxon")

resid_sex + resid_taxon
```

<center>**The independence condition is** **satisfied**. The lemurs could reasonably be treated as independent.</center>

## What conditions are needed? 

+--------------------+-----------------------+----------------------------+-------------------------+
| Model condition    | Interpret and predict | Simulation-based inference | Theory-based inference  |
+====================+=======================+============================+=========================+
| **Linearity**      | Important             | Important                  | Important               |
+--------------------+-----------------------+----------------------------+-------------------------+
| **Independence**   | Important             | Important                  | Important               |
+--------------------+-----------------------+----------------------------+-------------------------+
| **Normality**      | Not needed            | Not needed                 | Not needed if $n >> 30$ |
|                    |                       |                            |                         |
|                    |                       |                            | Important if $n < 30$   |
+--------------------+-----------------------+----------------------------+-------------------------+
| **Equal variance** | Not needed            | Not needed                 | Important               |
+--------------------+-----------------------+----------------------------+-------------------------+

# Model diagnostics

## Model diagnostics

```{r}
lemurs_aug <- augment(lemurs_fit)

lemurs_aug |> slice(1:10)
```

## Model diagnostics in R {.midi}

Use the `augment()` function in the broom package to output the model diagnostics (along with the predicted values and residuals)

-   response and predictor variables in the model
-   `.fitted`: predicted values
-   `.se.fit`: standard errors of predicted values
-   `.resid`: residuals
-   `.hat`: leverage
-   `.sigma`: estimate of residual standard deviation when the corresponding observation is dropped from model
-   `.cooksd`: Cook's distance
-   `.std.resid`: standardized residuals

## Influential Point

An observation is **influential** if removing has a noticeable impact on the regression coefficients

```{r,echo=F}
set.seed(12)
n <- 20
x <- c(runif(n,0,1))
y <- 3*x + rnorm(n,0,.5)
new.pt <- data.frame(x=2,y=0)
x.new <- c(x,2)
y.new <- c(y,0)
data <- bind_cols(x=x.new,y=y.new)
p1<- ggplot(data=data,aes(x=x,y=y))+geom_point(alpha =0.5)  + 
              geom_point(data=new.pt,color="red",size=3) + 
  geom_smooth(method="lm",se=F) + 
  labs(title = "With Influential Point")+ theme_light()+
  theme(title=element_text(hjust=0.5,size=14)) + 
  scale_x_continuous(limits = c(0,2)) 

data2 <- bind_cols(x=x,y=y)
p2 <- ggplot(data=data2,aes(x=x,y=y))+geom_point(alpha=0.5) + geom_smooth(method="lm",se=F) + 
  labs(title="Without Influential Point") + 
  scale_x_continuous(limits = c(0, 2)) + theme_light() + theme(title=element_text(hjust=0.5,size=14))  
p1 + p2
```

## Influential points

::: incremental
-   Influential points have a noticeable impact on the coefficients and standard errors used for inference
-   These points can sometimes be identified in a scatterplot if there is only one predictor variable
    -   This is often not the case when there are multiple predictors
-   We will use measures to quantify an individual observation's influence on the regression model
    -   **leverage**, **standardized & studentized residuals**, and **Cook's distance**
:::

# Leverage

## Leverage {.midi}

::: incremental
-   **Leverage** $(h_i)$ : a measure of the distance of the predictor (or combination of predictors) for the $i^{th}$ observation is from the mean value of the predictor (or combination of predictors)

-   For simple linear regression:

$$
h_i = \frac{1}{n} + \frac{(x_i - \bar{x})^2}{\sum_{j = 1}^n(x_j - \bar{x})^2}
$$

-   Observations with large values of $h_{i}$ are far away from the typical value (or combination of values) of the <u>predictors</u> in the data
:::

::: aside
See [Appendix A.4](https://intro-regression.github.io/appendix-linear.html#sec-hat-matrix) for math details of leverage in multiple linear regression.
:::

## Large leverage

::: incremental
-   The sum of the leverages for all points is $p + 1$, where $p$ is the number of predictors in the model

-   The average value of leverage is $\bar{h} =  \frac{(p+1)}{n}$

-   An observation has **large leverage** if $$h_{i} > \frac{2(p+1)}{n}$$
:::

## Lemurs: Leverage

```{r}
h_threshold <- 2 * (5 + 1) / nrow(lemurs)
h_threshold
```

. . .

```{r}
lemurs_aug |>
  filter(.hat > h_threshold)
```

## Let's look at the data

```{r}
#| echo: false

ggplot(data = lemurs, aes(x = litter_size)) + 
  geom_histogram() + 
  facet_wrap(~ taxon) +
  labs(x = "Litter size",
       title = "Litter size by taxon")
```

**The high leverage points are primarily VRUB lemurs with no siblings.**

## Large leverage

If there is point with high leverage, ask

-   ❓ Is there a data entry error?

-   ❓ Is this observation within the scope of individuals for which you want to make predictions and draw conclusions?

-   ❓ Is this observation impacting the estimates of the model coefficients? (Need more information!)

. . .

Just because a point has high leverage does not necessarily mean it will have a substantial impact on the regression. Therefore we need to check other measures.

# Scaled residuals

## Scaled residuals

::: incremental
-   What is the best way to identify outlier points that don't fit the pattern from the regression line?

    -   Look for points that have large residuals

-   We can rescale residuals and put them on a common scale to more easily identify "large" residuals

-   We will consider two types of scaled residuals: standardized residuals and studentized residuals
:::

## Standardized residuals

::: incremental
-   The variance of the residuals can be estimated by the mean squared residuals (MSR) $= \frac{SSR}{n - p - 1} = \hat{\sigma}^2_{\epsilon}$

-   We can use MSR to compute **standardized residuals**

    $$
    std.res_i = \frac{e_i}{\hat{\sigma}_{\epsilon}}
    $$

-   Standardized residuals are produced by `augment()` in the column `.std.resid`
:::

------------------------------------------------------------------------

## Using standardized residuals

We can examine the standardized residuals directly from the output from the `augment()` function

```{r}
#| echo: false
#| label: std residuals vs fitted
#| fig-align: center


ggplot(data = lemurs_aug, aes(x = .fitted, y = .std.resid)) + 
  geom_point() +
  geom_hline(yintercept = 3, color = "red", linetype =2)+
  geom_hline(yintercept = -3, color = "red", linetype = 2) +
  labs(x  = "Fitted values",
       y = "Std. Residuals",
       title = "Standardized residuals vs. fitted")
```

-   An observation is a *potential outlier* if its standardized residual is beyond $\pm 3$

## Digging in to the data

Let's look at the value of the response variable to better understand potential outliers

```{r}
#| echo: false
lemurs_aug %>%
  mutate(large_resid = if_else(abs(.std.resid) > 3, "Yes", "No"),
         large_resid = factor(large_resid, levels = c("Yes", "No")),
         obs_num = row_number()) |>
  ggplot(aes(x = obs_num, y = weight)) +
  geom_point(aes(color = large_resid), size = 2) + 
  labs(x = "Observation number",
    y = "Weight (in grams)", title = "Weight of PCOQ lemurs", 
    color = "Large std. residual") +
  #theme(legend.position = "none") + 
  scale_color_viridis(discrete = T, end = 0.8)
```

## Studentized residuals {.midi}

::: incremental
-   MSR is an approximation of the variance of the residuals.

-   A more precise calculation of the variance for the $i^{th}$ residual is is $Var(e_i) = \hat{\sigma}^2_{\epsilon}(1 - h_{i})$

    -   This is called the **studentized residual**
:::

$$
r_i = \frac{e_{i}}{\sqrt{\hat{\sigma}^2_{\epsilon}(1 - h_{i})}}
$$

-   Standardized and studentized residuals provide similar information about which points are outliers in the response.
    -   **Studentized residuals are used to compute Cook's Distance**

# Cook's Distance

## Motivating Cook's Distance

-   An observation's influence on the regression line depends on

    -   How close it lies to the general trend of the data

    -   Its leverage

-   **Cook's Distance** is a statistic that includes both of these components to measure an observation's overall impact on the model

## Cook's Distance {.midi}

Cook's distance for the $i^{th}$ observation is

$$
D_i = \frac{r^2_i}{p + 1}\Big(\frac{h_{i}}{1 - h_{i}}\Big)
$$

where $r_i$ is the studentized residual and $h_{i}$ is the leverage for the $i^{th}$ observation

. . .

This measure is a combination of

-   How well the model fits the $i^{th}$ observation (magnitude of residuals)

-   How far the combination of predictors for the $i^{th}$ observation is from the rest of the observations

## Using Cook's Distance

-   An observation with large value of $D_i$ is said to have a strong influence on the predicted values

-   General thresholds .An observation with

    -   $D_i > 0.5$ is **moderately influential**

    -   $D_i > 1$ is **very influential**

## Cook's Distance

Cook's Distance is in the column `.cooksd` in the output from the `augment()` function

```{r}
#| echo: false

lemurs_aug |>
  mutate(obs_num = row_number()) |>
  ggplot(aes(x = obs_num, y = .cooksd)) + 
  geom_point() + 
  geom_hline(yintercept = 1,color = "red") +
  labs(x = "Observation number",
       y = "Cook's distance")
```

## Using these measures

-   Standardized residuals, leverage, and Cook's Distance should all be examined together

-   Examine plots of the measures to identify observations that are outliers, high leverage, and may potentially impact the model.

## What to do with outliers/influential points?

::: incremental
-   First consider if the outlier is a result of a data entry error.

-   If not, you may consider dropping an observation if it's an outlier in the <u>predictor</u> variables if...

    -   It is meaningful to drop the observation given the context of the problem

    -   You intended to build a model on a smaller range of the predictor variables. Mention this in the write up of the results and be careful to avoid extrapolation when making predictions
:::

## What to do with outliers/influential points?

::: incremental
-   It is generally **not** good practice to drop observations that ar outliers in the value of the <u>response</u> variable

    -   These are legitimate observations and should be in the model

    -   You can try transformations or increasing the sample size by collecting more data

-   A general strategy when there are influential points is to fit the model with and without the influential points and compare the outcomes
:::

## Recap

-   Introduced model conditions for linear regression

-   Defined influential points

-   Introduced model diagnostics for linear regression

    -   Leverage

    -   Studentized residuals

    -   Cook's Distance

## Next class 

-   Exam 01 review

-   No prepare assignment
