---
title: "Multicollinearity"
author: "Prof. Maria Tackett"
date: "02-24-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 12-multicollinearity-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Announcements

-   Project proposal due **Thursday** t 11:59pm

-   Adeli's office hours next week: Monday, 9:50-11:40am in Old Chem 203B

## Announcements - Events

-   February 25, 5-6 pm \@ East Duke 108: Duke Sports Analytics Club is hosting **Sam Marks**, Director of Business Strategy and Analytics at the Boston Bruins

    -   [Click here](https://docs.google.com/forms/d/e/1FAIpQLSdethCogomC5DTNwSyQy5N_R5fOkT5LfJGR27VSEHGfWIH-cw/viewform) to register

-   SSMU Mini DataFest #2: February 27 - March 1. [Click here](https://edstem.org/us/courses/91299/discussion/7720540) for announcement

-   March 20 - 22: ASA DataFest at Duke

    -   Learn more and register: <https://dukestatsci.github.io/datafest/#signup> 

## Computing set up

```{r}
#| echo: true
#| message: false

# load packages
library(tidyverse)  
library(tidymodels)  
library(knitr)       
library(patchwork)
library(GGally)   # for pairwise plot matrix
library(corrplot) # for correlation matrix

# set default theme in ggplot2
ggplot2::theme_set(ggplot2::theme_bw())
```

## Topics

-   Multicollinearity

    -   Definition

    -   How it impacts the model

    -   How to detect it

    -   What to do about it

## Data: Trail users {.midi}

-   The Pioneer Valley Planning Commission (PVPC) collected data at the beginning a trail in Florence, MA for ninety days from April 5, 2005 to November 15, 2005
-   Data collectors set up a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.

```{r}
#| echo: false
#| message: false
#| warning: false

rail_trail <- read_csv(here::here("slides", "data/rail_trail.csv"))
rail_trail |> slice(1:5)
```

Source: [Pioneer Valley Planning Commission](http://www.fvgreenway.org/pdfs/Northampton-Bikepath-Volume-Counts%20_05_LTA.pdf) via the **mosaicData** package.

## Variables

**Response**

-   `volume` estimated number of trail users that day (number of breaks recorded)

**Predictors**

-   `hightemp` daily high temperature (in degrees Fahrenheit)

-   `avgtemp` average of daily low and daily high temperature (in degrees Fahrenheit)

-   `season` one of “Fall”, “Spring”, or “Summer”

-   `precip` measure of precipitation (in inches)

## EDA: Relationship between predictors

We can create a pairwise plot matrix using the `ggpairs` function from the **GGally** R package

```{r}
#| eval: false

rail_trail |>
  select(hightemp, avgtemp, season, precip) |>
  ggpairs()
```

## EDA: Relationship between predictors {.midi}

```{r}
#| echo: false
#| fig-align: center

rail_trail |>
  select(hightemp, avgtemp, season, precip) |>
  ggpairs()
```

## EDA: Correlation matrix

We can. use `corrplot()` in the **corrplot** R package to make a matrix of pairwise correlations between [quantitative]{.underline} predictors

```{r}
#| eval: false
correlations <- rail_trail |>
  select(hightemp, avgtemp, precip) |>
  cor()

corrplot(correlations, method = "number")
```

## EDA: Correlation matrix

```{r}
#| echo: false
#| fig-align: center

correlations <- rail_trail |>
  select(hightemp, avgtemp, precip) |>
  cor()


corrplot(correlations, method = "number")
```

## Question 

::::: panel-tabset
## Question

:::: midi
::: question
What is a potential concern about a model that uses high temperature, average temperature, season, and precipitation to predict volume?
:::

🔗 <https://forms.office.com/r/tPPLSe8eVg>
::::

## Submit

```{=html}
<center><iframe width="640px" height="480px" src="https://forms.office.com/r/tPPLSe8eVg?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe></center>
```
:::::

# Multicollinearity

## Multicollinearity {.midi}

::: incremental
-   Ideally the predictors are completely independent of one another

-   In practice, there is typically some relationship between predictors but it is often not a major issue in the model

-   If there predictors are perfectly correlated, we cannot find values of $\hat{\beta}_0, \hat{\beta}_1, \ldots, \hat{\beta}_p$ that best fit the model

-   If predictors are strongly correlated, we can find $\hat{\beta}_0, \hat{\beta}_1, \ldots, \hat{\beta}_p$, but there may be other issues with the model

-   **Multicollinearity**: predictors are strongly correlated with each other
:::

::: aside
Source: @montgomery2021introduction
:::

## Sources of multicollinearity

::: incremental
-   Data collection method - only sample from a subspace of the region of predictors

-   Constraints in the population - e.g., predictors family income and size of house

-   Choice of model - e.g., adding high order or interaction terms to the model

-   Overdefined model - have more predictors than observations
:::

## Example: Issue with multicollinearity

Let's assume the true population regression equation is $y = 3 + 4x$

. . .

Suppose we try estimating that equation using a model with variables $x$ and $z = x/10$

$$
\begin{aligned}\hat{y}&= \hat{\beta}_0 + \hat{\beta}_1x  + \hat{\beta}_2z\\
&= \hat{\beta}_0 + \hat{\beta}_1x  + \hat{\beta}_2\frac{x}{10}\\
&= \hat{\beta}_0 + \bigg(\hat{\beta}_1 + \frac{\hat{\beta}_2}{10}\bigg)x
\end{aligned}
$$

## Example: Issue with mulitcollinearity

$$\hat{y} = \hat{\beta}_0 + \bigg(\hat{\beta}_1 + \frac{\hat{\beta}_2}{10}\bigg)x$$

-   We can set $\hat{\beta}_1$ and $\hat{\beta}_2$ to any two numbers such that $\hat{\beta}_1 + \frac{\hat{\beta}_2}{10} = 4$

-   Therefore, we are unable to choose the "best" combination of $\hat{\beta}_1$ and $\hat{\beta}_2$

## Variance inflation factor {.midi}

-   The **variance inflation factor (VIF)** is a measure of the collinearity between predictor $x_j$ and all other predictors in the model

$$
VIF_{j} = \frac{1}{1 - R^2_j}
$$

where $R^2_j$ is the proportion of variation in $x_j$ that is explained by all the other predictors

## Detecting multicollinearity

-   Common practice uses threshold $VIF > 10$ as indication of concerning multicollinearity (some say VIF \> 5 is worth investigation)

-   Variables with similar values of VIF are typically the ones correlated with each other

-   Use the `vif()` function in the **rms** R package to calculate VIF

```{r}
library(rms)

trail_fit <- lm(volume ~ hightemp + avgtemp + precip, data = rail_trail)

vif(trail_fit)
```

# Application exercise

::: appex
📋 [sta210-sp26.github.io/ae/ae-08-multicollinearity.html](../ae/ae-08-multicollinearity.html)

Part 1
:::

## How multicollinearity impacts model {.midi}

::: incremental
-   When we have perfect collinearities, we are unable to get estimates for the coefficients

-   When we have almost perfect collinearities (i.e. highly correlated predictor variables), the standard errors for our regression coefficients inflate

    -   In other words, we lose precision in our estimates of the regression coefficients

    -   This impedes our ability to use the model for inference

-   It is also difficult to interpret the model coefficients
:::

## Dealing with multicollinearity

::: incremental
-   Collect more data (often not feasible given practical constraints)

-   Redefine the correlated predictors to keep the information from predictors but eliminate collinearity

    -   e.g., if $x_1, x_2, x_3$ are correlated, use a new variable $(x_1 + x_2) / x_3$ in the model

-   For categorical predictors, avoid using levels with very few observations as the baseline

-   Remove one of the correlated variables

    -   Be careful about substantially reducing predictive power of the model
:::

# Application exercise

::: appex
📋 [sta210-sp26.github.io/ae/ae-08-multicollinearity.html](../ae/ae-08-multicollinearity.html)

Part 2
:::

## Recap

-   Introduced multicollinearity

    -   Definition

    -   How it impacts the model

    -   How to detect it

    -   What to do about it

## Next class

-   Variable transformations

-   Complete [Lecture 13 prepare](../prepare-lec13.html)

## References
