---
title: "Logistic regression"
subtitle: "Inference"
author: "Prof. Maria Tackett"
date: "03-24-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 18-logistic-inference-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Project EDA due TODAY at 11:59pm (on GitHub)

-   HW 04 due March 21 at 11:59pm

    -   Released after class

-   Statistics experience due April 15

-   SSMU Data Mini #3 - April 4

    -   See Ed Discussion for full announcement

## Topics

-   Model assessment using ROC curve and AUC

-   Estimating coefficients for logistic regression model

-   Inference for a single model coefficient

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(openintro)
library(knitr)
library(kableExtra)  # for table embellishments
library(patchwork)

# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))
```

## Data: Risk of coronary heart disease

This data set is from an ongoing cardiovascular study on residents of the town of Framingham, Massachusetts.

-   `high_risk`: 1 = High risk of having heart disease in next 10 years, 0 = Not high risk of having heart disease in next 10 years

-   `age`: Age at exam time (in years)

-   `totChol`: Total cholesterol (in mg/dL)

-   `currentSmoker`: 0 = nonsmoker; 1 = smoker

**The goal is to use logistic regression to identify patients who are high risk of heart disease.**

## Univariate EDA

```{r}
#| echo: false
heart_disease <- read_csv("data/framingham.csv") |>
  select(age, totChol, TenYearCHD, currentSmoker) |>
  drop_na() |>
  mutate(high_risk = as_factor(TenYearCHD), 
         currentSmoker = as_factor(currentSmoker))
```

```{r}
#| echo: false

p1 <- ggplot(data = heart_disease, aes(x = high_risk)) + 
  geom_bar(fill = "darkcyan", color = 'black') + 
  labs(x = "High risk")

p2 <- ggplot(data = heart_disease, aes(x = age)) + 
  geom_histogram(fill = "darkcyan",color = 'black', binwidth = 2) + 
  labs(x = "Age")
p3 <- ggplot(data = heart_disease, aes(x = totChol)) + 
  geom_histogram(fill = "darkcyan", color = 'black') + 
  labs(x = "Total Cholesterol")

p4 <- ggplot(data = heart_disease, aes(x = currentSmoker)) + 
  geom_bar(fill = "darkcyan", color = 'black') + 
  labs(x = "Current smoker")

(p1 + p2) / (p3 + p4)
```

## Bivariate EDA

```{r}
#| echo: false

p5 <- ggplot(data = heart_disease, aes(x = high_risk, y = age)) + 
  geom_boxplot(fill = "darkcyan") +
  labs(x = "High risk",
       y = "Age") + 
  coord_flip()

p6 <- ggplot(data = heart_disease, aes(x = currentSmoker, fill = high_risk)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "Current smoker", 
       fill = "High risk") + 
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Dark2")

p7 <- ggplot(data = heart_disease, aes(x = high_risk, y = totChol)) + 
  geom_boxplot(fill = "steelblue") +
  labs(x = "High risk",
       y = "Total cholesterol") + 
  coord_flip()

p6 + (p5 / p7) 

```

## Risk of coronary heart disease

```{r}
#| echo: true

heart_disease_fit <- glm(high_risk ~ age + totChol + currentSmoker, 
              data = heart_disease, family = "binomial")
```

```{r}
#| echo: false

tidy(heart_disease_fit) |> 
  kable(digits = 3)
```

# ROC curve and AUC

## Prediction and classification

::: incremental
-   We are often interested in using the model to classify observations, i.e., predict whether a given observation will have a 1 or 0 response

-   For each observation

    -   Use the logistic regression model to calculate the predicted log-odds the response for the $i^{th}$ observation is 1
    -   Use the log-odds to calculate the predicted probability the $i^{th}$ observation is 1
    -   Then, use the predicted probability to classify the observation as having a 1 or 0 response using some predefined threshold
:::

## Confusion matrix

A **confusion matrix** is a $2 \times 2$ table that compares the predicted and actual classes for a given threshold. We can produce this matrix using the `conf_mat()` function in the **yardstick** package (part of tidymodels).

<br>

```{r}
#| echo: false

heart_disease_aug <- augment(heart_disease_fit, type.predict = "response")

heart_disease_aug <- heart_disease_aug |>
  mutate(pred_class = factor(if_else(.fitted > 0.5, 1, 0)))

heart_conf_mat <- heart_disease_aug |>
  conf_mat(high_risk, pred_class)

autoplot(heart_conf_mat, type = "heatmap") 
```

## True/false positive/negative {.midi}

|   | <u>Not</u> high risk $(y_i = 0)$ | High risk $(y_i = 1)$ |
|------------------------|------------------------|------------------------|
| **Classified** <u>**not**</u> **high risk** $(\hat{\pi}_i \leq \text{threshold})$ | True negative (TN) | False negative (FN) |
| **Classified high risk** $(\hat{\pi}_i > \text{threshold})$ | False positive (FP) |  |

## Sensitivity and specificity

::::: panel-tabset
## Question

:::: midi
::: question
1.  How do we set the threshold if the primary goal is high sensitivity?

2.  How do we set the threshold if the primary goal is high specificity?
:::

🔗 <https://forms.office.com/r/ysNdXsj5cW>
::::

## Submit

```{=html}
<center><iframe width="640px" height="480px" src="https://forms.office.com/r/ysNdXsj5cW?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe></center>
```
:::::

## ROC Curve {.midi}

So far the model assessment has depended on the model and selected threshold. The **receiver operating characteristic (ROC) curve** allows us to assess the model performance across a range of thresholds.

. . .

::::: columns
::: {.column width="65%"}
```{r}
#| echo: false

# calculate sensitivity and specificity at each threshold
roc_curve_data <- heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second") 

# plot roc curve
autoplot(roc_curve_data)
```
:::

::: {.column width="35%"}
-   x-axis: 1 - Specificity (False positive rate)

-   y-axis: Sensitivity (True positive rate)

Which corner of the plot indicates the best model performance?
:::
:::::

## ROC curve

```{r}
#| echo: false
#| out-width: 100%

heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second")  |>
  autoplot() +
  annotate("point", x = 0, y = 1, color = "#5B888C") +
  annotate("point", x = 0, y = 1, color = "#5B888C", size = 3, shape = "circle open") +
  annotate(
    "label", x = 0.02, y = 0.99, label = "All high risk classified as high risk,\nall not high risk classified as not high risk", hjust = 0, color = "#5B888C", fontface = "bold", vjust = 1, fill = "white"
  ) +
  annotate("point", x = 1, y = 0, color = "#8F2D56") +
  annotate("point", x = 1, y = 0, color = "#8F2D56", size = 3, shape = "circle open") +
  annotate(
    "label", x = 0.98, y = 0.01, label = "All high risk classified as not high risk,\nall not high risk classified as high risk", hjust = 1, color = "#8F2D56", fontface = "bold", vjust = 0, fill = "white"
  ) +
  annotate(
    "segment", color = "#5b708c", x = 0, y = 0, xend = 1, yend = 1, size = 2, alpha = 0.6
  ) +
  annotate(
    "label", x = 0.58, y = 0.5, label = "True positive rate\n= false positive rate", hjust = 0, color = "#5b708c", fontface = "bold", fill = "white"
  )
```

## ROC curve in R {.midi}

```{r}
#| echo: true
#| eval: false


# calculate sensitivity and specificity at each threshold
roc_curve_data <- heart_disease_aug |>
  roc_curve(high_risk, .fitted, 
            event_level = "second") 

# plot roc curve
autoplot(roc_curve_data)
```

## ROC curve in R

::::: columns
::: {.column width="40%"}
Randomly selected observations from `roc_data`

```{r}
#| echo: false

set.seed(210)

roc_curve_data |>
  sample_n(20)

```
:::

::: {.column width="60%"}
```{r}
#| echo: false

autoplot(roc_curve_data)
```
:::
:::::

## Area under the curve

The **area under the curve (AUC)** can be used to assess how well the logistic model fits the data

-   AUC=0.5: model is a very bad fit (no better than a coin flip)

-   AUC close to 1: model is a good fit

. . .

```{r}
#| echo: true
heart_disease_aug |>
  roc_auc(high_risk, .fitted,
    event_level = "second"
  )
```

# Estimating coefficients

## Statistical model {.midi}

The form of the statistical model for logistic regression is

$$
\log\Big(\frac{\pi_i}{1-\pi_i}\Big) = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2} + \dots + \beta_px_{ip}
$$

where $\pi_i$ is the probability $y_i = 1$.

## Distribution of $Y$

The response variable $Y$ follows a Bernoulli distribution, $\text{Bernoulli}(\pi)$ (similarly we can think of $n$ observed $Y$'s as generated from a Binomial distribution, $\text{Binomial}(n,\pi)$)

. . .

For an individual observation,

$$
p(y_i | x_{i1}, \ldots, x_{ip}, \beta_0, \beta_1, \ldots, \beta_p) = \pi_i^{y_i}(1 - \pi_i)^{1- y_i}
$$

## Likelihood function

The likelihood function is the joint density of $y_1, \ldots, y_n$ for unknown coefficients $\beta_0, \ldots, \beta_p$

$$
\begin{aligned}
L &= \prod_{i=1}^{n}
p(y_i | x_{i1}, \ldots, x_{ip}, \beta_0, \beta_1, \ldots, \beta_p) \\[8pt]
& = \prod_{i=1}^{n}
\pi_i^{y_i}(1 - \pi_i)^{1- y_i}
\end{aligned}
$$

. . .

The estimated coefficients are the combination of $\hat{\beta}_0, \hat{\beta}_1, \ldots, \hat{\beta}_p$ that maximize the likelihood $L$. They are the **maximum likelihood estimates**.

## Estimating coefficients {.midi}

We use the log-likelihood function to find the maximum likelihood estimates

$$
\log L = \sum\limits_{i=1}^n[y_i \log(\hat{\pi}_i) + (1 - y_i)\log(1 - \hat{\pi}_i)]
$$

where

$\hat{\pi} = \frac{\exp\{\hat{\beta}_0 + \hat{\beta}_1X_1 + \dots + \hat{\beta}_pX_p\}}{1 + \exp\{\hat{\beta}_0 + \hat{\beta}_1X_1 + \dots + \hat{\beta}_pX_p\}}$

## Iterative model fitting in R

```{r}
#| include: false

trace(glm.fit, quote(cat <- function(...) {
  base::cat(...)
  if (...length() >= 3 && identical(..3, " Iterations - ")) print(coefold)
}))
```

```{r}
high_risk_fit_full <- glm(high_risk ~ age + currentSmoker,
                          data = heart_disease, family = "binomial", 
                          # print each iteration
                        control = glm.control(trace = TRUE))

#stop printing for future models
untrace(glm.fit)
```

# Inference for coefficients

## Inference for coefficients

There are two approaches for testing coefficients in logistic regression

-   **(Wald) hypothesis test.** Use to test

    -   a single coefficient (when $n$ is large)

<!-- -->

-   **Drop-in-deviance test**. Use to test...

    -   a single coefficient
    -   a categorical predictor with 3+ levels
    -   a group of predictor variables

## Hypothesis test for $\beta_j$

**Hypotheses:** $H_0: \beta_j = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_j \neq 0$, given the other variables in the model

. . .

**(Wald) Test Statistic:** $$z = \frac{\hat{\beta}_j - 0}{SE(\hat{\beta}_j)}$$

. . .

**P-value:** $P(|Z| > |z|)$, where $Z \sim N(0, 1)$, the Standard Normal distribution

## Confidence interval for $\beta_j$

We can calculate the **C% confidence interval** for $\beta_j$ as the following:

$$
\hat{\beta}_j \pm z^* \times SE_{\hat{\beta}_j}
$$

where $z^*$ is calculated from the $N(0,1)$ distribution

. . .

::: callout-note
This is an interval for the change in the log-odds for every one unit increase in $x_j$
:::

## Interpretation in terms of the odds

The change in **odds** for every one unit increase in $x_j$.

$$
\exp\{\hat{\beta}_j \pm z^* \times SE_{\hat{\beta}_j}\}
$$

. . .

**Interpretation:** We are $C\%$ confident that for every one unit increase in $x_j$, the odds multiply by a factor of $\exp\{\hat{\beta}_j - z^*\times SE_{\hat{\beta}_j}\}$ to $\exp\{\hat{\beta}_j + z^* \times SE_{\hat{\beta}_j}\}$, holding all else constant.

## Coefficient for `age` {.midi}

```{r}
#| label: risk-model-age-highlight
#| code-fold: true

high_risk_fit <- glm(high_risk ~ age +  currentSmoker,
                          data = heart_disease, family = "binomial",
                     control = glm.control(trace = FALSE))

tidy(high_risk_fit, conf.int = TRUE) |> 
  kable(digits = 5) |>
  row_spec(2, background = "#D9E3E4")
```

. . .

**Hypotheses:**

$$
H_0: \beta_{age} = 0 \hspace{2mm} \text{ vs } \hspace{2mm} H_a: \beta_{age} \neq 0
$$, given current smoking status is in the model

## Coefficient for `age`    {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Test statistic:**

$$z = \frac{0.08360 - 0}{0.00556} = 15.04
$$

## Coefficient for `age`  {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**P-value:**

$$
P(|Z| > |15.04|) \approx 0
$$

. . .

```{r}
2 * pnorm(15.04,lower.tail = FALSE)
```

## Coefficient for `age`  {.midi}

```{r}
#| echo: false
#| ref.label: risk-model-age-highlight
```

**Conclusion:**

The p-value is very small, so we reject $H_0$. The data provide sufficient evidence that age is a statistically significant predictor of whether someone is high risk of having heart disease, after accounting for smoking status.

## CI for `age`

```{r}
#| echo: false

tidy(high_risk_fit, conf.int = TRUE) |> 
  kable(digits = 3) |>
  row_spec(2, background = "#D9E3E4")
```

::: question
Interpret the 95% confidence interval for `age` in terms of the **odds** of being high risk for heart disease.
:::

## Recap

-   Model assessment using ROC curve and AUC

-   Estimating coefficients for logistic regression model

-   Inference for a single model coefficient

## Next class

-   Model selection for logistic regression

-   Complete [Lecture 19 prepare](../prepare/prepare-lec19.hthml)
