---
title: "Special topic: Causal inference"
author: "Prof. Maria Tackett"
date: "04-02-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 21-causal-inference-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Project:

    -   Preliminary analysis due April 7

    -   Presentations April 14 & 16

-   Statistics experience due April 15

-   Exam 02 April 9 (in-class), April 9 - 11 (take-home)

    -   Lecture recordings + practice questions available on website

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)
library(MatchIt) # for propensity score matching
library(ggridges)
library(patchwork)


# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))

# set color palettes 
sunset2 <- PNWColors::pnw_palette("Sunset2",2)
sunset3 <- PNWColors::pnw_palette("Sunset2",3)
sunset4 <- PNWColors::pnw_palette("Sunset2",4)

```

## Data science ethics case study

A data analyst received permission to analyze a data set that was scraped from a social media site. The full data set included name, screen name, email address, geographic location, IP (internet protocol) address, demographic profiles, and preferences for relationships.

<br>

::: question
What are ethical considerations of putting a deidentified data set with name and email address removed in a LLM (e.g., Claude or ChatGPT) to help with analysis?
:::

::: aside
Adapted from Chapter 8 of @baumer2017modern
:::

# Causal inference

## Introduce scenario

```{r}
project_ace <- read_csv("https://intro-regression.github.io/data/project-ace-data.csv") |>
  mutate(Tracking.Pathway = factor(if_else(is.na(Tracking.Pathway), "Control", "Treatment")), 
         Grade = factor(Grade))
```

## Look at distribution of GPA

```{r}
ggplot(data = project_ace, aes(x = Current.GPA, y = Tracking.Pathway, fill = Tracking.Pathway)) + 
  geom_density_ridges() + 
  labs(x = "Grade Point Average (GPA)", 
       y = "Treatment/ Control") + 
  theme_bw() + 
  theme(legend.position = "none") + 
  scale_fill_manual(values = sunset2)
```

::: question
How does the distribution of GPA compare between the two groups?
:::

## EDA: Covariates by group


```{r}
#| echo: false

# grade level 

p_grade <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = Grade)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion", 
       fill = "Grade") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset4)

# gender

p_gender <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = Gender)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

# ELL

p_ell <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = ELL)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
   theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

p_grade + p_gender + p_ell
```

## Try a model

```{r}

orig_model <- lm(Current.GPA ~ Grade + Gender + Ethnicity.Hispanic.Y.N + ELL + Sped + Homeless, data = project_ace)

tidy(orig_model) |> 
  kable(digits = 3)
```


-   Why cant we draw causal claims about the effect of the treatment?

## Two types of data 

-   Experimental

-   Observational

## Drawing conclusions from observational data

## Propensity score matching 

-   What it is and the process
- how to choose what goes in the propensity score model

## Propensity score model 

```{r}
#| echo: true
propensity_score_model <- glm(Tracking.Pathway ~ Grade + Gender + Ethnicity.Hispanic.Y.N + ELL + Sped + Homeless,
                        data = project_ace, 
                        family = "binomial")
```

```{r}
tidy(propensity_score_model) |> 
  kable(digits = 3)
```

## Common support 

```{r}
#| echo: true

# compute propensity scores
project_ace_aug <- augment(propensity_score_model, 
                           type.predict = "response")
```


```{r}
ggplot(data = project_ace_aug, aes(x = .fitted, y = Tracking.Pathway, fill = Tracking.Pathway)) + 
  geom_density_ridges() +
  labs(x = "Propensity score",
       y = "Treatment/ Control") +
  theme_bw() + 
  theme(legend.position = "none") + 
  scale_fill_manual(values = sunset2)
```

- What do you observe about these distributions?

-   Why do we want these distributions to overlap?

## How matching happens

-   is there a visual to help with this?

## Match it R package + code

## Look at matched results 

## Look at the matched data set

## Fit the treatment model 

## Average treatment effect 

-   What is your conclusion about project ACE?

## Recap

# Poll question 

What is one question / topic you want to cover on the Exam 02 review. Exam 02 will cover topics from Multicollinearity - Today.

## Further reading

## Next class

-   Exam 02 review

-   No prepare assignment

## References
