---
title: "Causal inference"
author: "Prof. Maria Tackett"
date: "04-02-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
  html: 
    output-file: 21-causal-inference-notes.html
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| include: false

# figure options
knitr::opts_chunk$set(
  fig.width = 8, fig.asp = 0.618, out.width = "90%",
  fig.retina = 3, dpi = 300, fig.align = "center"
)

library(countdown)
```

## Announcements

-   Project:

    -   Preliminary analysis due April 7

    -   Presentations April 14 & 16

-   Statistics experience due April 15

-   Exam 02 April 9 (in-class), April 9 - 11 (take-home)

    -   Lecture recordings + practice questions available on website

## Computational setup

```{r}
#| echo: true

# load packages
library(tidyverse)
library(tidymodels)
library(knitr)
library(MatchIt) # for propensity score matching
library(ggridges)
library(patchwork)


# set default theme and larger font size for ggplot2
ggplot2::theme_set(ggplot2::theme_bw(base_size = 20))

# set color palettes 
sunset2 <- PNWColors::pnw_palette("Sunset2",2)
sunset3 <- PNWColors::pnw_palette("Sunset2",3)
sunset4 <- PNWColors::pnw_palette("Sunset2",4)

```

## Data science ethics case study

A data analyst received permission to analyze a data set that was scraped from a social media site. The full data set included name, screen name, email address, geographic location, IP (internet protocol) address, demographic profiles, and preferences for relationships.

<br>

::: question
What are ethical considerations of putting a deidentified data set with name and email address removed in a LLM (e.g., Claude or ChatGPT) to help with analysis?
:::

::: aside
Adapted from Chapter 8 of @baumer2017modern
:::

# Causal inference

## Impact of Project ACE

Project Action for Equity (Project ACE) was a "five year interdisciplinary program aimed to get more underrepresented high school students from disadvantaged backgrounds to get interested in college degrees in engineering as well as biomedical and behavioral sciences" [@ProjectACE_UTEP].

**Our goal is toevaluate whether the data provide evidence that participation in Project ACE improved students' educational outcomes (as measured by GPA).**

The data were obtained from @evans2025testing, and the analysis will closely follow the analysis in their article.

```{r}
project_ace <- read_csv("https://intro-regression.github.io/data/project-ace-data.csv") |>
  mutate(Tracking.Pathway = factor(if_else(is.na(Tracking.Pathway), "Control", "Treatment")), 
         Grade = factor(Grade))
```

## Variables {.midi}

The data are in `project-ace-data.csv`. We will use the following variables:

-   `Grade`: Grade level (9, 10, 11, 12)
-   `Gender`: Gender (F, M)
-   `Ethnicity`: Ethnicity (Hispanic, Non-Hispanic)
-   `ELL`: Whether student is an English language learner (N, Y)
-   `Sped`: Whether student is in a special education program (N, Y)
-   `Homeless`: Whether student is homeless (N, Y)
-   `Tracking.Pathway`: Whether student was in Project ACE (Treatment) or not (Control)
-   `Current.GPA`: Grade Point Average (GPA) ranging 0 to 5.0

## Glimpse of data

```{r}
#| echo: true
project_ace <- read_csv("https://intro-regression.github.io/data/project-ace-data.csv") |>
  mutate(Tracking.Pathway = factor(if_else(is.na(Tracking.Pathway), "Control", "Treatment")), 
         Grade = factor(Grade))
```

<br>

```{r}
#| echo: true


glimpse(project_ace)
```

## Look at distribution of GPA

```{r}
ggplot(data = project_ace, aes(x = Current.GPA, y = Tracking.Pathway, fill = Tracking.Pathway)) + 
  geom_density_ridges() + 
  labs(x = "Grade Point Average (GPA)", 
       y = "Treatment/ Control") + 
  theme_bw() + 
  theme(legend.position = "none") + 
  scale_fill_manual(values = sunset2)
```

::: question
How does the distribution of GPA compare between the two groups?
:::

## Covariates by group 

```{r}
#| echo: false

# grade level 

p_grade <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = Grade)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion", 
       fill = "Grade") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset4)

# gender

p_gender <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = Gender)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

# ELL

p_ell <- ggplot(data = project_ace, aes(x = Tracking.Pathway, fill = ELL)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
   theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

p_grade + p_gender + p_ell
```

## Try a model {.midi}

```{r}
orig_model <- lm(Current.GPA ~ Grade + Gender + Ethnicity.Hispanic.Y.N + ELL + Sped + Homeless, data = project_ace)

tidy(orig_model) |> 
  kable(digits = 3)
```

## Draw causal conclusion? 

::: question
Why should we avoid using this model to conclude participation in Project ACE improved students' GPAs?
:::

## Two types of data

-   **Experimental data**: Data obtained by explicitly applying a treatment

    -   Individuals randomly assigned to treatment or control groups
    -   Can draw causal conclusions, because we can assume the only difference between treatment and control groups is the treatment itself

-   **Observational data**: Data obtained without explicitly applying a treatment

    -   Potential underlying factors that are confounded with response likelihood of being in treatment group

    -   Cannot draw causal conclusions from typical regression analysis

## Causal conclusions from observational data

-   We will use statistical methods to make observational data look more like experimental data

-   We do so by making the treatment and control groups similar based on a set of covariates (variables) that impact both the response and the likelihood an individual is in the treatment group

-   We create these groups matching individuals in the treatment and control groups based on these underlying covariates

    -   We do this matching using the process called **propensity score matching**

## Propensity score matching 

In **propensity score matching**, we compute the probability (propensity) an observation is assigned to the treatment group based on a set of confounding variables that directly impact both the response and the likelihood an individual is in the treatment group. We then match individuals in the treatment and control groups who have the same or similar propensities to create a new data set in which the control and treatment groups have similar characteristics. The matched data set is used to estimate the effect of the treatment. <!--# use exchangeability here?-->

-   What it is and the process
-   how to choose what goes in the propensity score model

## Propensity score model

```{r}
#| echo: true
propensity_score_model <- glm(Tracking.Pathway ~ Grade + Gender + Ethnicity.Hispanic.Y.N + ELL + Sped + Homeless,
                        data = project_ace, 
                        family = "binomial")
```

```{r}
tidy(propensity_score_model) |> 
  kable(digits = 3)
```

## Common support

**Common support**: Individuals in the matched treatment and control groups have a non-zero probability of receiving the treatment

```{r}
#| echo: true

# compute propensity scores
project_ace_aug <- augment(propensity_score_model, 
                           type.predict = "response")
```

```{r}
ggplot(data = project_ace_aug, aes(x = .fitted, y = Tracking.Pathway, fill = Tracking.Pathway)) + 
  geom_density_ridges() +
  labs(x = "Propensity score",
       y = "Treatment/ Control") +
  theme_bw() + 
  theme(legend.position = "none") + 
  scale_fill_manual(values = sunset2)
```

::: question
-   What do you observe from the visualization?

-   Why is common support important?
:::

## How matching happens

-   is there a visual to help with this?

## Propensity score matching in R

We conduct propensity score matching using `matchit()` from the **MatchIt** R package (will need to install in the containers).

```{r}
#| echo: true
#| label: match-scores-nearest-neighbor
#| 
library(MatchIt)

# generate propensity scores and matches
project_ace_psm <- matchit(Tracking.Pathway ~ Grade + Gender + Ethnicity.Hispanic.Y.N +
                          Race + ELL + Sped + Homeless,
                 data = project_ace,
                 method = "nearest",   
                 distance = "logit")

# data with matches
project_ace_matched <- match.data(project_ace_psm)
```

-   `method = "nearest"`: Each observation in the treatment group is matched the an observation in the control group with closest propensity score

-   `distance = "logit"`: Generate the propensity scores using a logistic regression model

## Matched data

```{r}
project_ace_matched |> slice(1:10)
```

## Matched observations

```{r}
set.seed(123)

groups <- sample(1:148, 3, replace = FALSE)

project_ace_matched |> 
  filter(subclass %in% groups) |>
  arrange(subclass, Tracking.Pathway) |>
  select(subclass, Tracking.Pathway, distance, Tracking.Pathway, Grade, Gender, Ethnicity.Hispanic.Y.N, ELL, Sped, Homeless) |>
  kable()
```

## Covariates by group in matched data

```{r}
#| echo: false

# grade level 

p_grade <- ggplot(data = project_ace_matched, aes(x = Tracking.Pathway, fill = Grade)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion", 
       fill = "Grade") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset4)

# gender

p_gender <- ggplot(data = project_ace_matched, aes(x = Tracking.Pathway, fill = Gender)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
  theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

# ELL

p_ell <- ggplot(data = project_ace_matched, aes(x = Tracking.Pathway, fill = ELL)) + 
  geom_bar(position = "fill", color = "black") +
  labs(x = "", 
       y = "Proportion") +
   theme_bw() + 
  theme(legend.position = "bottom") +
  scale_fill_manual(values = sunset2)

p_grade + p_gender + p_ell
```

## Fit the treatment model

```{r}
#| echo: true
#| 
treatment_model <- lm(Current.GPA ~ Tracking.Pathway, 
                      data = project_ace_matched)
```

```{r}
tidy(treatment_model) |> 
  kable(digits = 3, conf.int = TRUE)
```

<br>

::: question
Describe the effect of Project ACE on GPA. Is there evidence that participating in the program leads higher GPA?
:::

## Average treatment effect

-   What is your conclusion about project ACE?

## Recap

## Further reading 

All books are freely available online.

-   [Causal Inference: The Mixtape](https://mixtape.scunning.com/) by Scott Cunningham

-   [The Effect: An Introduction to Research Design](https://theeffectbook.net/) and Causality by Nick Huntington-Klein

-   [Causal Inference in R](https://www.r-causal.org/) by Malcolm Barret, Lucy McGowan, Travis Gerke

## Exam 02 review

We will do Exam 02 review in lab on Monday, April 6 and lecture on Tuesday, April 7. Exam 02 covers content from multicollinearity (February 24) - today.

::::: panel-tabset
## Question

:::: midi
::: question
Please submit one question you have about the Exam 02 content. I will use these questions to write the exam reviews.
:::

🔗 <https://forms.cloud.microsoft/r/0ukue0JaMw>
::::

## Submit

```{=html}
<center><iframe width="640px" height="480px" src="https://forms.cloud.microsoft/r/0ukue0JaMw?embed=true" frameborder="0" marginwidth="0" marginheight="0" style="border: none; max-width:100%; max-height:100vh" allowfullscreen webkitallowfullscreen mozallowfullscreen msallowfullscreen> </iframe></center>
```
:::::

## Next class

-   Exam 02 review

-   No prepare assignment

## References
