---
title: "Lab 01"
subtitle: "Linear regression"
date: "01-26-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

## Tips for working on lab

-   You do not have to finish the lab in class, they will always be due **Thursday**. One work strategy is to get through portions that you think will be most challenging (which initially might be the coding component) during lab when a TA can help you on the spot and leave the narrative writing until later.

-   Do not pressure each other to finish early (particularly once you start working on teams); use the time wisely to really learn the material and produce a quality report.

## Tips: Axis labels and titles

::: midi
Below is a graph of association between flipper length in millimeters and body mass in grams of three species of penguins in Palmer Station, Antarctica. What are informative title and axis labels for this graph?
:::

```{r}
#| echo: false
#| fig-align: center
#| out-width: "60%"

library(palmerpenguins)
library(tidyverse)
library(viridis)

ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) + 
  geom_point(size = 2) + 
  theme_bw() +
    scale_color_manual(values = c("darkorange","purple","cyan4")) 
```

## Tips: Code style

Which code chunk would you rather read?

```{r}
#| eval: false
# code chunk 1
penguins|>filter(!is.na(flipper_length_mm))|>group_by(species)|>summarise(min=min(flipper_length_mm),mean=mean(flipper_length_mm),sd=sd(flipper_length_mm),max=max(flipper_length_mm),n=n())
```

<br>

. . .

```{r}
#| eval: false
# code chunk 2
penguins |> 
  filter(!is.na(flipper_length_mm)) |> 
  group_by(species) |> 
  summarise(min = min(flipper_length_mm), 
            mean = mean(flipper_length_mm), 
            max = max(flipper_length_mm),
            n = n())
```

## Tips: Code style cont'd

Make code easier to read and debug by

-   Putting each element on a different line (start a new line after `+` and `|>`)

-   Putting spaces before and after operators (`+`, `-`, `*`, `=`, `|>` )

-   In general, avoiding long lines of code, i.e. lines longer than 80 characters.

See the [R for Data Science - Workflow: code style](https://r4ds.hadley.nz/workflow-style.html) for more tips on code styling.

## **When you're done with lab**

-   Make sure all your final changes have been pushed to your GitHub repo
-   Submit the PDF of your responses to [**Gradescope**](https://www.gradescope.com/)

## Lab 01: Linear regression

-   Today's lab focuses on using linear regression to explore the relationship between air temperature and ice duration for two lakes in Wisconsin, along with the change in ice duration over time.

-   There are markers throughout suggesting when to render, commit, and push changes to GitHub. These are to help you start using version control in your workflow.

-   There are points for having a neatly formatted document and implementing a reproducible workflow

🔗 [sta210-sp26.github.io/labs/lab-01.html](https://sta210-sp26.github.io/labs/lab-01.html)
