---
title: "Lab 03"
subtitle: "Multiple linear regression"
date: "02-09-2026"
date-format: "MMM DD, YYYY"
footer: "🔗 [sta210-sp26.github.io](https://sta210-sp26.github.io)"
logo: "../images/logo.png"
format: 
  revealjs: 
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    include-before: [ '<script type="text/x-mathjax-config">MathJax.Hub.Config({tex2jax: {enableAssistiveMml: false}});</script>']
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
execute:
  freeze: auto
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r}
#| echo: false

library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
```

## Goals

-   Practice: Inference for multiple linear regression
-   Team workflow
-   Lab 03: Candy competition

# Inference for multiple linear regression

## Data: NCAA Football expenditures {.midi}

Today's data come from [Equity in Athletics Data Analysis](https://ope.ed.gov/athletics/#/datafile/list) and includes information about sports expenditures and revenues for colleges and universities in the United States. This data set was featured in a [March 2022 Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-29/readme.md).

We will focus on the 2019 - 2020 season expenditures on football for 127 institutions in the NCAA - Division 1 FBS. The variables are :

-   `total_exp_m`: Total expenditures on football in the 2019 - 2020 academic year (in millions USD)

-   `enrollment_th`: Total student enrollment in the 2019 - 2020 academic year (in thousands)

-   `type`: institution type (Public or Private)

```{r}
football <- read_csv("data/ncaa-football-exp.csv")
```

## Bivariate EDA

```{r}
#| code-fold: true
p1 <- ggplot(data = football, aes(x = enrollment_th, y = total_exp_m)) +
  geom_point() + 
  labs(x = "Enrollment (in thousands)", 
       y = "Total expenditures (in millions)", 
       title = "Football expenditures vs. enrollment")

p2 <- ggplot(data = football, aes(x = type, y = total_exp_m)) +
  geom_boxplot() + 
  labs(x = "Institution type", 
       y = "Total expenditures (in millions)", 
       title = "Football expenditures vs. institution type")

p1 + p2
```

## Potential interaction effect?

```{r}
#| code-fold: true
ggplot(data = football, aes(x = enrollment_th, y = total_exp_m, color = type,
                            linetype = type)) +
  geom_point(alpha = 0.7) + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Enrollment (in thousands)", 
       y = "Total expenditures (in millions)", 
       title = "Football expenditures vs. enrollment",
       subtitle = "by institution type", 
       color = "Type", 
       linetype = "Type") +
  theme_bw()
```

## Inference {.midi}

```{r}
#| code-fold: true
football_fit <- lm(total_exp_m ~ enrollment_th + type + enrollment_th * type, 
                   data = football)

tidy(football_fit) |> 
  kable(digits = 3)
```

::: question
Let's look at the test for `enrollment_th:typePublic`.

-   State the null and alternative hypotheses being tested in the model output.

-   Explain what `std.error = 0.387` quantifies.

-   Explain how the test statistic `2.364` is computed and what it means in the context of the data.
:::

## Inference {.midi}

```{r}
#| code-fold: true
football_fit <- lm(total_exp_m ~ enrollment_th + type + enrollment_th * type, 
                   data = football)

tidy(football_fit) |> 
  kable(digits = 3)
```

::: question
-   State the distribution used to compute the p-value.

-   The coefficient of `enrollment_th:typePublic` has a low p-value but the coefficient for `enrollment_th` has a high p-value. What does this mean about the relationship between football expenditures and enrollment?
:::

## Meet your team!

-   [**Click here**](https://prodduke-my.sharepoint.com/:x:/g/personal/mt324_duke_edu/IQBOyCiOAbuCTaSDqSwx5e17AXQXY39VvMy2szPYTCGCMUY?e=muOdca) to find your team for this week's lab.

-   Sit with your team.

## Team workflow

-   Only <u>**one**</u> team member should type at a time. There are markers in today's lab to help you determine whose turn it is to type.

    -   Every team member should still be engaged in discussion for all questions, even if it's not your turn type.

-   Don't forget to **pull** to get your teammates' updates before making changes to the `.qmd` file.

    ::: callout-important
    **Only one submission per team on Gradescope. Read the submission instructions carefully!**
    :::

## Team workflow, in action

-   Complete the "Workflow: Using Git and GitHub as a team" section of the lab in your teams.
-   Raise your hand if you have any questions about the workflow.

```{r}
#| echo: false
countdown::countdown(minutes = 5, font_size = "2em")
```

## Tips for working on a team {.midi}

-   Do not pressure each other to finish early; use the time wisely to really learn the material and produce a quality report.

-   The labs are structured to help you learn the steps of a data analysis. Do not split up the lab among the team members; work on it together in its entirety.

-   Everyone has something to contribute! Use the lab groups as an opportunity to share ideas and learn from each other.

## Today's lab

In today's lab, your team will compare multiple models to find the best one to describe what makes a good candy.

🔗 <https://sta210-sp26.github.io/labs/lab-03.html>
