---
title: "AE 07: Exam 01 review"
subtitle: "Trail users"
date: "February 12, 2026"
editor: visual
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-sp26) and locate your `ae-07` repo to get started.

Render, commit, and push your responses to GitHub by the end of class.
:::

## Packages

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(tidymodels)
library(knitr)
library(patchwork)
```

## Trail users

The Pioneer Valley Planning Commission (PVPC) collected data for ninety days from April 5, 2005 to November 15, 2005. Data collectors set up a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.

We will use regression analysis to predict the number of trail users based on weather and other features describing the day.

We'll use the following variables in this analysis:

-   `volume` estimated number of trail users that day (number of breaks recorded)

-   `hightemp` daily high temperature (in degrees Fahrenheit)

-   `daytype` one of "weekday" or "weekend"

View the data set[^1] to see the remaining variables.

[^1]: Source: [Pioneer Valley Planning Commission](http://www.fvgreenway.org/pdfs/Northampton-Bikepath-Volume-Counts%20_05_LTA.pdf) via the **mosaicData** package.

```{r}
#| message: false

rail_trail <- read_csv("data/rail-trail.csv")
```

## Bivariate EDA

```{r}
p1 <- ggplot(data = rail_trail, aes(x = hightemp, y = volume)) + 
  geom_point() + 
  labs(x = "High temperature (in Fahrenheit)", 
       y = "Number of users", 
       title = "Volume versus high temperature")

p2 <- ggplot(data = rail_trail, aes(x = daytype, y = volume)) + 
  geom_boxplot() + 
  labs(x = "", 
       y = "Number of users", 
       title = "Volume versus day type")

p1 + p2
```

## Models

Below is the code for the two models used in this analysis.

```{r}
# main effects model 
main_effects_model <- lm(volume ~ hightemp + daytype, data = rail_trail)


# interaction model 
interaction_model <- lm(volume ~ hightemp + daytype + hightemp * daytype,
                         data = rail_trail)

```

## Exercises

### Exercise 1

Consider the **main effects** model.

-   Display the model along with the 90% confidence intervals for the coefficients.

<!-- -->

-   Interpret the coefficient of `hightemp` in the context of the data.

-   Interpret the coefficient of `daytypeWeekend` in the context of the data.

-   Does it make sense to interpret the intercept? Explain your reasoning.

-   If not, what can we do to make the interpretation meaningful?

### Exercise 2

Consider the **interactions effects** model.

-   Display the model along with the 90% confidence intervals for the coefficients.

<!-- -->

-   Interpret the main effect of `hightemp`.

-   Interpret the coefficient `hightemp:daytypeWeekend` .

-   According to this model, does the effect of `hightemp` differ for weekends vs. weekdays? Explain.

### Exercise 3

The following code can be used to create a bootstrap distribution for the model coefficients in the **main effects model**.

-   Describe what each line of code does.

```{r}
#| label: ex-6

set.seed(210)

boot_dist <- rail_trail |> # <1>
  specify(volume ~ hightemp + daytype) |> # <2>
  generate(reps = 100, type = "bootstrap") |> # <3>
  fit() # <4>


```

1.  \_\_\_

2.  \_\_\_

3.  \_\_\_

4.  \_\_\_

-   How many observations are in each bootstrap sample?

-   Explain why we use bootstrap sampling, instead of permutation sampling, to construct confidence intervals.

### Exercise 4

Consider the bootstrap distribution for the coefficient of `hightemp` in the **main effects** model.

-   How many estimated coefficients of `hightemp` make up the bootstrap distribution?

-   Run the code from the previous exercise to construct the bootstrap distribution. Use the bootstrap distribution to compute $SE_{\hat{\beta}_\text{hightemp}}$. Explain what this value means in the context of the data.

-   Construct a 90% bootstrap confidence interval for the coefficient of `hightemp`. Interpret the interval in the context of the data.

### Exercise 5

We would like to conduct the following hypothesis test for the **main effects** model.

$$
\begin{aligned}
&H_0: \beta_\text{hightemp} = 0 \\
&H_a: \beta_\text{hightemp} \neq 0
\end{aligned}
$$

-   State the hypotheses in words.

-   Conduct a permutation test by permuting the response variable. Use 100 iterations, a seed of 210, and a significance level of $\alpha = 0.05$.

### Exercise 6

-   Use the **interaction effects** model to predict the volume for a Saturday that has a high temperature of 75 degrees Fahrenheit.

-   Suppose you're asked to construct a confidence interval and a prediction interval given a Saturday with a high temperature of 75 degrees Fahrenheit. Which interval would you expect to be wider and why? In your answer clearly state the difference between these intervals.

-   Construct the intervals and interpret them in the context of the data.

# Wrapping up

::: callout-important
Once you’ve completed the AE:

-   Render the document to produce the PDF with all of your work from today’s class.

-   Push all your work to your AE repo on GitHub. You’re done! 🎉
:::
