---
title: "AE 06: Inference for multiple linear regression"
subtitle: "Palmer penguins"
date: "February 5, 2026"
editor: visual
---

::: callout-important
Go to the [course GitHub organization](https://github.com/sta210-sp26) and locate your `ae-06` repo to get started.
:::

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)
library(knitr)
```

## Data: Palmer penguins

The `penguins` data set contains data for penguins found on three islands in the Palmer Archipelago, Antarctica. Data were collected and made available by [Dr. Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) and the [Palmer Station, Antarctica LTER](https://pal.lternet.edu/), a member of the [Long Term Ecological Research Network](https://lternet.edu/). These data can be found in the **palmerpenguins** R package.

```{r}
#| echo: true

penguins <- palmerpenguins::penguins |>
  select(body_mass_g, flipper_length_mm, bill_length_mm, species) |>
  drop_na()
```

::: callout-caution
For this analysis, we will drop observations that have missing values for any of the variables of interest. In practice, we want to be careful about how this might bias our conclusions.
:::

## Main effects model

Let's fit the main effects model using `flipper_length_mm`, `species`, and `bill_length_mm`.

```{r}
#| label: main-effects-model

penguins_fit <- lm(body_mass_g ~ flipper_length_mm +
                     species + bill_length_mm, 
                   data = penguins)

tidy(penguins_fit, conf.int = TRUE) |>
  kable(digits = 3)
```

## Bootstrap CI

### Construct the bootstrap CI

Fill in the code to construct a bootstrap distribution for `speciesGentoo`. Remove `#| eval: false` after filling in the code.

```{r}
#| label: construct-ci
#| eval: false

n = 100
set.seed(210)

boot_dist <- penguins |>
  specify(______) |>
  generate(reps = n, type = "bootstrap") |>
  fit()
```

### Confidence interval for `speciesGentoo`

Use the bootstrap distribution to compute the standard error for `speciesGentoo`. How does it compare to the standard error in the model output? Is this what you expected? Why or why not?

```{r}
#| label: se-beta

# add code

```

Construct the 95% confidence interval for `speciesGentoo`. How does it compare to the 95% CI in the model output? Is this what you expect? Why or why not?

```{r}
#| label: bootstrap-ci

# add code
```

## Permutation test

Similar to simple linear regression, we can use permutation sampling for simulation-based hypothesis tests for coefficients in multiple linear regression.

Now that we have multiple predictors, there are many ways to permute the data. We will discuss two here:

-   Permute the response variable.

-   Permute the predictor of interest.

We will use both methods to conduct a hypothesis test for the coefficient of `speciesGentoo`.

### Permute response variable

Fill in the code to construct permutation samples by permuting the response variable for the variable in [Main effects model]. We will also compute the p-value for the coefficient. Remove `#| eval: false` after you have filled in the code.

```{r}
#| label: permute-response-var
#| eval: false

n = 100
set.seed(210)

# construct null distribution 
null_dist <- _____ |>
  specify(______) |>
  hypothesize(null = "independence") |>
  generate(reps = _____, type = "permute") |>
  fit()

# observed fit
observed_fit <- penguins |>
  specify(body_mass_g ~ flipper_length_mm +
                     species + bill_length_mm) |>
  fit()

# compute the p-value
get_p_value(
  ____,
  obs_stat = ____,
  direction = "two-sided"
)
```

What is your conclusion about the coefficient of `speciesGentoo` based on the p-value?

### Permute single predictor

Now let's construct the permutation sample by permuting only the variable `species`. We do so by adding the `variables` argument to the generate function. Remove `#| eval: false` after you have filled in the code.

```{r}
#| label: permute-single-var
#| eval: false

n = 100
set.seed(210)

# construct null distribution 
null_dist_2 <- _____ |>
  specify(______) |>
  hypothesize(null = "independence") |>
  generate(reps = _____, type = "permute", 
           variables = c(species)) |>
  fit()

# observed fit
observed_fit <- penguins |>
  specify(body_mass_g ~ flipper_length_mm +
                     species + bill_length_mm) |>
  fit()

# compute the p-value
get_p_value(
  ____,
  obs_stat = ____,
  direction = "two-sided"
)
```

What is your conclusion about the coefficient of `speciesGentoo` based on the p-value?

### Different conclusions

Why did we get different conclusions? *Hint: Think about what structure / variable relationships are preserved or broken in each approach.*

# Wrapping up

::: callout-important
Once you’ve completed the AE:

-   Render the document to produce the PDF with all of your work from today’s class.

-   Push all your work to your AE repo on GitHub. You’re done! 🎉
:::
