---
title: "AE 12: Exam 02 Review"
date: "April 7, 2026"
format: 
  html:
    code-fold: true
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the course GitHub organization and locate your `ae-12` repo to get started
:::

## Packages

```{r}
#| label: load-pkgs
 
library(tidyverse)
library(tidymodels)
library(knitr)
library(pROC)
```

# Part 1: Multiple linear regression

```{r}
insurance <- read_csv('data/insurance.csv')

# sample down 18 - 19 year olds
set.seed(210)
insurance_young <- insurance |>
  filter(age == 18 | age == 19) |>
  sample_n(50)

insurance <-  insurance |>
  filter(age > 19) |>
  bind_rows(insurance_young) |>
  mutate(log_charges = log(charges),
         age_cent = age - mean(age),
         region = case_when(region == "northeast" ~ "NorthEast", 
                            region =="northwest" ~ "NorthWest", 
                            region == "southeast" ~ "SouthEast",
                            region == "southwest" ~ "SouthWest")
  )
```

## Data: Insurance charges

The **insurance** data set contains information about patient characteristics and the total charges billed by a medical insurance company. There are `r nrow(insurance)` observations in the data set. We will use the following variables:

-   `age_cent`: Mean-centered age in years (the mean age in the data is `r round(mean(insurance$age), 1)` years old)

-   `children`: Number of children covered by the insurance policy

-   `smoker`: Indicator of whether the the patient smokes regularly (1: yes, 0: no)

-   `region`: Region where the patient lives (NorthEast, SouthEast, NorthWest, SouthWest )

-   `charges`: Total costs billed by the medical insurance company

***The goal of the analysis is to use age, number of children covered by the insurance policy, smoking status, and region to predict the total costs billed by the medical insurance company. The log-transformed value of charges,*** `log(charges)`***, is used to fit the model.***

```{}
```

## Exercise 1

Why do we use `log(charges)` instead of the original response variable in this analysis?

## Exercise 2

The output of the main effects model is below.

```{r}
model1 <- lm(log_charges ~ age_cent + children + smoker + region, data = insurance)

tidy(model1) |>
  kable(digits = 3)
```

-   Interpret `age_cent` in the context of the data in terms of `charges` (not `log(charges)`)

-   Interpret `regionSouthEast` in the context of the data in terms of `charges` (not `log(charges)`).

## Exercise 3

We want to compare the model from the previous exercise to a model that also includes the interaction between `age_cent` and `smoker`. To do so, we begin by splitting the data into training and testing sets.

The code to split the data into training and testing data is below. Explain why we may want to split the data into training and testing sets.

```{r}
set.seed(210)

insurance_split <- initial_split(insurance, prop = 0.8)
insurance_train <- training(insurance_split)
insurance_test <- testing(insurance_split)
```

## Exercise 4

We will use 5-fold cross validation to determine which model is the best fit. Explain why we want to use cross validation to compare models in this analysis.

## Exercise 5

The code for cross validation is below. Describe the process of cross validation. You can use the code as a guide.

Then, state which model you select based on the results of cross validation.

```{r}
set.seed(210)


folds <- vfold_cv(insurance_train, v = 5)

# main effects model
main_effects_workflow <- workflow() |> 
  add_model(linear_reg()) |> 
  add_formula(log_charges ~ age_cent + children + smoker + region) 

main_effects_cv <- main_effects_workflow |> 
  fit_resamples(resamples = folds)

collect_metrics(main_effects_cv, summarize = TRUE)

# interaction effects model
interaction_effects_workflow <- workflow() |> 
  add_model(linear_reg()) |> 
  add_formula(log_charges ~ age_cent + children + smoker + region + age_cent * smoker) 

interaction_effects_cv <- interaction_effects_workflow |> 
  fit_resamples(resamples = folds)

collect_metrics(interaction_effects_cv, summarize = TRUE)
```

# Part 2: Logistic regression

## Data: Social media ads

```{r}
ads <- read_csv("data/Social_Network_Ads.csv")
# https://www.kaggle.com/datasets/rakeshrau/social-network-ads

# sample to have even purchases / not purchases

set.seed(210)
#no_purchase <- ads |>
#  filter(Purchased == 0) |>
#  sample_n(143)

# Make purchased a factor
# Split age in categories 
# Rescale salary by 1000s
ads <- ads |>
 # filter(Purchased == 1) |>
 # bind_rows(no_purchase) |>
  mutate(purchased = factor(Purchased),
         salary = EstimatedSalary/1000,
         age_category = case_when(
           Age < 35 ~ "18-34",
           Age >= 35 & Age < 50 ~ "35-49", 
           TRUE ~ "50+"
         ))
```

The **ads** data set contains information about users on a social media website and whether they purchased a product through an advertisement (ad) on the site. There are `r nrow(ads)` observations in the data set. We will use the following variables:

-   `salary`: Individual's estimated salary (in thousands of dollars)

-   `age_category`: Category indicating an individual's age in years (18-34, 35-49, 50+)

-   `purchased`: Whether an individual purchased a product through an ad on the social media website (1: yes, 0: no)

***The goal of the analysis is to predict the odds an individual purchases a product through an on the social media website based on their estimated salary and age.***

## Exercise 6

We begin by fitting a model using `age_category` to predict the odds an individual purchases a product through a social media ad. The output is below.

```{r}
ad_age_fit <- glm(purchased ~ age_category, data = ads,
                  family = "binomial")

tidy(ad_age_fit) |>
  kable(digits = 3)
```

-   What is the odds ratio of `35-49` versus `18-34`?

-   What are the odds an 52 year old purchases a product from a social media website? What is the probability?

## Exercise 7

Now let's fit the model that includes both `salary` and `age_category`. The output is below.

```{r}
ad_age_salary_fit <- glm(purchased ~ age_category + salary, data = ads,
                  family = "binomial")

tidy(ad_age_salary_fit) |>
  kable(digits = 3)
```

-   What is the adjusted odds ratio of `35-49` versus `18-34`?

-   What are the predicted odds of a 52 year old who has a salary of \$60,000?

-   What is the predicted probability for this individual?

## Exercise 8

The ROC curve for the fitted model is below.

```{r}
ad_age_salary_aug <- augment(ad_age_salary_fit, type.predict = "response")

ad_age_salary_aug |>
  roc_curve(
    purchased,
    .fitted,
    event_level = "second"
) |>
  autoplot()
```

-   Explain what each point on the curve represents.

-   The code to compute AUC is below. Use AUC to evaluate model performance.

```{r}
ad_age_salary_aug |>
  roc_auc(
    purchased,
    .fitted,
    event_level = "second"
)
```

## Exercise 9

*General question.*

Suppose we are comparing models, such that different models are selected based on AIC and BIC.

-   When might we prefer the model selected by AIC?

-   When might we prefer the model selected by BIC?

# Submission

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today’s class.

-   Push all your work to your AE repo on GitHub. You’re done! 🎉
:::
