---
title: "AE 11: Model selection for logistic regression"
date: "March 26, 2026"
editor: visual
execute: 
  warning: false
  message: false
---

::: callout-important
Go to the course GitHub organization and locate your `ae-11` repo to get started.
:::

## Packages

```{r}
#| label: load-pkgs-data
#| message: false
 
library(tidyverse)
library(tidymodels)
library(knitr)

leukemia <- read_csv("data/leukemia.csv") |>
  mutate(Resp = factor(Resp))
```

## Response to Leukemia treatment

Today's data is from a study where 51 untreated adult patients with Acute Myeloid Leukemia who were given a course of treatment, and they were assessed as to their response to the treatment.[^1]

[^1]: The data set is from the Stat2Data R package. This AE is adapted from exercises in *Stat 2.*

The goal of today's analysis is to use pre-treatment factors to predict how likely it is a patient will respond to the treatment.

We will use the following variables:

-   `Age`: Age at diagnosis (in years)
-   `Smear`: Differential percentage of blasts
-   `Infil`: Percentage of absolute marrow leukemia infiltrate
-   `Index`: Percentage labeling index of the bone marrow leukemia cells
-   `Blasts`: Absolute number of blasts, in thousands
-   `Temp`: Highest temperature of the patient prior to treatment, in degrees Fahrenheit
-   `Resp`: 1 = responded to treatment or 0 = failed to respond


## Exercise 1

Begin by splitting the data into training (80%) and testing (20%) sets. Use `210` for the seed.




## Exercise 2

We will begin by considering the model with all pre-treatment variables: `Age`, `Smear`, `Infil`, `Index`, `Blasts` and `Temp`. Fit a model using these six variables to predict whether a patient responded to the treatment (`Resp`).

Which variables are statistically significant using a threshold of $\alpha = 0.05$?




## Exercise 3

We want to compare the model from Exercise 2 that includes all pre-treatment predictors to a model that only includes the statistically significant predictors. We will conduct cross validation using AUC on the assessment set as the criteria for selecting a final model.

Why might we prefer cross-validation with AUC over AIC/BIC on the training data for this analysis?

## Exercise 4

Complete the code below to conduct 4-fold cross validation on the training data for the model in Exercise 2 with all pre-treatment predictors. Use `210` for the seed.
:::

::: callout-important
Remove `#| eval: false` when you have filled in the code.
:::

```{r}
#| label: ex-04
#| eval: false

# define the folds
folds <- vfold_cv(____, v = 4)

# specify workflow for the model
______ <- workflow() |> 
  add_model(logistic_reg()) |> 
  add_formula(__________) 

# CV for model 
cross_validation <- _______ |> 
  fit_resamples(resamples = folds)

# compute performance statistics for the model
collect_metrics(cross_validation, summarize = TRUE)
```


## Exercise 5

Conduct cross validation for the model that only includes the predictors from Exercise 2 that are statistically significant at $\alpha = 0.05$. Use `210` for the seed.


## Exercise 6

Which model do you select based on the results from cross validation? Evaluate how well the selected model performs on the testing data.


## Wrapping up

::: callout-important
Render the document to produce the PDF with all of your work from today's class.

Push all your work to your AE repo on GitHub. You're done! 🎉
:::
