---
title: "AE 10: Logistic regression"
date: "March 17, 2026"
editor: visual
execute:
  warning: false
  message: fase
---

::: callout-important
Go to the [course GitHub organization](github.com/sta210-sp26) and locate your `ae-10` repo to get started.
:::

```{r}
#| label: load-packages
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(tidymodels)
```

# Introduction

```{r}
#| echo: false
pew_data <- read_csv("data/pew-atp-w132.csv")
```

This data comes from the 2023 [Pew Research Center's American Trends Panel](https://www.pewresearch.org/the-american-trends-panel/). The survey aims to capture public opinion about a variety of topics including politics, religion, and technology, among others. We will use data from 11201 respondents in Wave 132 of the survey conducted July 31 - August 6, 2023.

A more complete analysis on this topic can be found in the Pew Research Center article [*Growing public concern about the role of artificial intelligence in daily life*](https://www.pewresearch.org/short-reads/2023/08/28/growing-public-concern-about-the-role-of-artificial-intelligence-in-daily-life/) by Alec Tyson and Emma Kikuchi.

**The goal of this analysis is to understand the relationship between age and concern about the increased use of AI in daily life.**

You will use the following variables:

-   `ai_concern`: Whether a respondent said they are "more concerned than excited" about in the increased use of AI in daily life (1: yes, 0: no)

-   `age_cat`: Age category

    -   18-29
    -   30-49
    -   50-64
    -   65+
    -   Refused

```{r}
#| label: load-prep-data
#| message: false
#| warning: false

pew_data <- pew_data |>
  mutate(ai_concern = if_else(CNCEXC_W132 == 2, 1, 0),
         age_cat = case_when(F_AGECAT == 1 ~ "18-29",
                             F_AGECAT == 2 ~ "30-49",
                             F_AGECAT == 3 ~ "50-64",
                             F_AGECAT == 4 ~ "65+", 
                             TRUE ~ "Refused"
))

# Make factors and  relevel 
pew_data <- pew_data |>
  mutate(ai_concern = factor(ai_concern),
         age_cat = factor(age_cat))
```

## Exercise 1

-   Compute the probability an 18-29 year-old respondent is concerned about AI.

-   Compute the **odds** an 18-29 year-old respondent is concerned about AI.

## Exercise 2

-   Compute the **odds** a 50-64 year-old respondent is concerned about AI.

-   How do the odds a 50-64 year-old respondent is concerned about AI compare to the odds an 18-29 year-old respondent is concerned? This is the odds ratio.

## Exercise 3

Fit the logistic regression model using `age_cat` to predict the log odds an individual is concerned about AI. Neatly display the model.

## Exercise 4

Use the model from the previous exercise to

-   What is the baseline level of `age_cat`?

-   Interpret the coefficient of `age_cat50-64` in terms of the log odds of being concerned about AI.

## Exercise 5

-   The model in Exercise 3 is in terms of the log odds. Describe how we can write the model in terms of the **odds** an individual is concerned about AI.

::: callout-tip
Recall the process we used for models with a log-transformed response variable.
:::

-   Interpret the coefficient of `age_cat50-64` in terms of the **odds** of being concerned about AI.

-   How does this interpretation compare to Exercise 2? What does this tell you about the coefficients in the logistic regression model?

# Wrapping up

::: callout-important
Once you’ve completed the AE:

-   Render the document to produce the PDF with all of your work from today’s class.

-   Push all your work to your AE repo on GitHub. You’re done! 🎉
:::
